---
title: "Google Data Analytics Capstone Project"
author: "Prashant Burman"
date: "31/03/2022"
output:
pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Case Study: How Does a Bike-Share Navigate Speedy Success?

The purpose of this document is to consolidate downloaded Divvy data into a single data frame and then conduct simple analysis to help answer the key question: In what ways do members and casual riders use Divvy bikes differently?

### Introduction
This exploratory analysis case study is towards Capstone project requirement for [Google Data Analytics Professional Certificate](https://www.coursera.org/professional-certificates/google-data-analytics). The case study involves a bikeshare company's data of its customer's trip details over a 12 month period (March 2021 - February 2022). The data has been made available by Motivate International Inc. under this [license](https://www.divvybikes.com/data-license-agreement).

The analysis will follow the 6 phases of the Data Analysis process: Ask, Prepare, Process, Analyze, and Act. A brief explanation of these processes:

#### Ask

- Ask effective questions
- Define the scope of the analysis
- Define what success looks like

#### Prepare

- Verify data’s integrity
- Check data credibility and reliability
- Check data types
- Merge datasets

#### Process

- Clean, Remove and Transform data
- Document cleaning processes and results


#### Analyze

- Identify patterns
- Draw conclusions
- Make predictions

#### Share

- Create effective visuals
- Create a story for data
- Share insights to stakeholders

#### Act

- Give recommendations based on insights
- Solve problems
- Create something new

<br/>

## 1. Ask


#### Scenario

Marketing team needs to design marketing strategies aimed at converting casual riders into annual members. In order to do that, however, the marketing analyst team needs to better understand how annual members and casual riders differ.

#### Stakeholders:

- Director of marketing
- Cyclistic executive team

##### Objective

Hence, the objective for this analysis is to throw some light on how the two types of customers: annual members and casual riders, use Cyclistic bikeshare differently, based on few parameters that can be calculated/ obtained from existing data.

#### Deliverables:

- Insights on how annual members and casual riders use Cyclistic bikes differently
- Provide effective visuals and relevant data to support insights
- Use insights to give three recommendations to convert casual riders to member riders

<br/>

## 2. Prepare

#### Data Source

A total of **12 CSV files** have been made available for each month starting from **March 2021 to February 2022**. Each file captures the details of every ride logged by the customers of Cyclistic. This data that has been made publicly available has been scrubbed to omit rider's personal information.

The combined size of all the 12 CSV files is close to 1 GB. Data cleaning in spreadsheets will be time-consuming and slow compared to R. I am choosing R simply because I could do both data wrangling and analysis/ visualizations in the same platform. 


<br/>

#### Load Libraries

```{r}

library(tidyverse)
library(ggplot2)
library(lubridate)
library(dplyr)
library(readr)
library(janitor)
library(data.table)
library(tidyr)
library(hms)
```
<br/>

#### Load original .csv files worth of data from March 2021 to Feb 2022
```{r}
setwd("C:/Users/prash/Downloads/DA/R markdown/New folder/Citibikeshare")

march21_df <- read_csv("202103-divvy-tripdata.csv")
april21_df <- read_csv("202104-divvy-tripdata.csv")
may21_df <- read_csv("202105-divvy-tripdata.csv")
june21_df <- read_csv("202106-divvy-tripdata.csv")
july21_df <- read_csv("202107-divvy-tripdata.csv")
aug21_df <- read_csv("202108-divvy-tripdata.csv")
sep21_df <- read_csv("202109-divvy-tripdata.csv")
oct21_df <- read_csv("202110-divvy-tripdata.csv")
nov21_df <- read_csv("202111-divvy-tripdata.csv")
dec21_df <- read_csv("202112-divvy-tripdata.csv")
jan22_df <- read_csv("202201-divvy-tripdata.csv")
feb22_df <- read_csv("202202-divvy-tripdata.csv")
```

## 3. Process


#### Merge all of the data frames into one year view

```{r}
cyclist_df <- rbind(march21_df,april21_df,may21_df,june21_df,july21_df,aug21_df,sep21_df,
oct21_df,nov21_df,dec21_df,jan22_df,feb22_df)
```

#### Remove individual month data frames to clear up space in the environment  
```{r}

remove(march21_df,april21_df,may21_df,june21_df,july21_df,aug21_df,sep21_df,
oct21_df,nov21_df,dec21_df,jan22_df,feb22_df)
```

#### Create new data frame to contain new columns
```{r}

cyclistic_date <- cyclist_df

```

#### started_at & ended_at should be in datetime datatype instead of char. Convert all from char to datetime.
```{r}
cyclistic_date[['started_at']] <- ymd_hms(cyclistic_date[['started_at']])
cyclistic_date[['ended_at']] <- ymd_hms(cyclistic_date[['ended_at']])
```
#### Rename Columns name for better readability
```{r}

names(cyclistic_date)[13] <- 'customer_type'
names(cyclistic_date)[2] <- 'ride_type'
```

#### Calculate ride length by subtracting ended_at time from started_at time and converted it to minutes
```{r}

cyclistic_date$ride_length <- difftime(cyclist_df$ended_at, cyclist_df$started_at, units = "mins")
```

#### Create columns for: day of week, month, day, year, time, hour
```{r}
cyclistic_date$date <- as.Date(cyclistic_date$started_at) #default format is yyyy-mm-dd, use start date

cyclistic_date$day_of_week <- wday(cyclist_df$started_at) #calculate the day of the week

cyclistic_date$day_of_week <- format(as.Date(cyclistic_date$date), "%A") #create column for day of week

cyclistic_date$month <- format(as.Date(cyclistic_date$date), "%m") #create column for month

cyclistic_date$day <- format(as.Date(cyclistic_date$date), "%d") #create column for day

cyclistic_date$year <- format(as.Date(cyclistic_date$date), "%Y") #create column for year

cyclistic_date$time <- format(as.Date(cyclistic_date$date), "%H:%M:%S") #format time as HH:MM:SS

cyclistic_date$time <- as_hms((cyclistic_date$started_at)) #create new column for time

cyclistic_date$hour <- hour(cyclistic_date$time) #create new column for hour
```

#### Create column for different seasons: Spring, Summer, Fall, Winter
```{r}

cyclistic_date <-cyclistic_date %>% mutate(season = 
                                             case_when(month == "03" ~ "Spring",
                                                       month == "04" ~ "Spring",
                                                       month == "05" ~ "Spring",
                                                       month == "06"  ~ "Summer",
                                                       month == "07"  ~ "Summer",
                                                       month == "08"  ~ "Summer",
                                                       month == "09" ~ "Fall",
                                                       month == "10" ~ "Fall",
                                                       month == "11" ~ "Fall",
                                                       month == "12" ~ "Winter",
                                                       month == "01" ~ "Winter",
                                                       month == "02" ~ "Winter"
                                                       ))
```
#### Clean the data
```{r}

cyclistic_date <- na.omit(cyclistic_date) #remove rows with NA values

cyclistic_date <- distinct(cyclistic_date) #remove duplicate rows 

cyclistic_date <- cyclistic_date[!(cyclistic_date$ride_length <=0),] #remove where ride_length is 0 or negative

cyclistic_date <- cyclistic_date <- cyclistic_date %>% #remove columns not needed
  select(-c( ride_id,start_station_id,end_station_id,start_lat,start_lng,end_lat,end_lng)) 
```
#### View the final data:
```{r}
glimpse(cyclistic_date)

head(cyclistic_date)
```
## 4&5. Analyze and Share Data:

### TOTAL RIDES
#### Total Number of Rides:
```{r}

nrow(cyclistic_date)
```

#### CUSTOMER TYPE:
```{r}

cyclistic_date %>%
  group_by(customer_type) %>% 
  count(customer_type)
```

#### TYPE OF BIKE:
Total Rides by customer_type 
```{r}
cyclistic_date %>%
  group_by(customer_type, ride_type) %>% 
  count(ride_type)
```
#### Total Rides: 
```{r}
cyclistic_date %>%
  group_by(ride_type) %>% 
  count(ride_type)
```  
  
### HOUR
#### Total Rides by customer_type: 
```{r}
cyclistic_date %>%
  group_by(customer_type) %>% 
  count(hour) %>% 
  print(n = 48) #lets you view the entire tibble
```

#### Total rides:
```{r}
cyclistic_date %>%
  count(hour) %>% 
  print(n = 24) #lets you view the entire tibble  
``` 
### DAY OF THE WEEK 

#### Total rides by customer_type:
```{r}
cyclistic_date %>%
  group_by(customer_type) %>% 
  count(day_of_week)
```
#### Total rides :
```{r}
cyclistic_date %>%
  count(day_of_week) 
```  
### DAY OF THE MONTH

#### Total rides by customer_type:
```{r}
cyclistic_date %>%
  group_by(customer_type) %>% 
  count(day) %>% 
  print(n = 62) #lets you view the entire tibble  
```
#### Total rides:
```{r}
cyclistic_date %>%
  count(day) %>% 
  print(n = 31) #lets you view the entire tibble
```  

### MONTH

#### Total rides by customer_type: 
```{r}
cyclistic_date %>%
  group_by(customer_type) %>% 
  count(month) %>% 
  print(n = 24) #lets you view the entire tibble
```
#### Total rides:
```{r}
cyclistic_date %>%
  count(month) 
```

### SEASON

#### Spring:
##### Total rides by customer_type:
```{r}
cyclistic_date %>%
  group_by(customer_type) %>% 
  filter(season == "Spring") %>% 
  count(season)
```
##### Total rides:
```{r}
cyclistic_date %>%
  filter(season == "Spring") %>% 
  count(season)
```
#### Summer:
##### Total rides by customer_type:
```{r}
cyclistic_date %>%
  group_by(customer_type) %>% 
  filter(season == "Summer") %>% 
  count(season)
```
##### Total rides:
```{r}
cyclistic_date %>%
  filter(season == "Summer") %>% 
  count(season)
```
#### Fall:
##### Total rides by customer_type:
```{r}
cyclistic_date %>%
  group_by(customer_type) %>% 
  filter(season == "Fall") %>% 
  count(season)
```
##### Total rides:
```{r}
cyclistic_date %>%
  filter(season == "Fall") %>% 
  count(season)
```
#### Winter:
##### Total rides by customer_type:
```{r}
cyclistic_date %>%
  group_by(customer_type) %>% 
  filter(season == "Winter") %>% 
  count(season)
```
##### Total rides: 
```{r}
cyclistic_date %>%
  filter(season == "Winter") %>% 
  count(season)
```
#### ALL SEASONS

##### Total rides by customer_type:
```{r}
cyclistic_date %>%
  group_by(season, customer_type) %>% 
  count(season)
```
##### Total rides:
```{r}
cyclistic_date %>%
  group_by(season) %>% 
  count(season)
```  
#### AVERAGE RIDE LENGTH

##### Average of ride_length:
```{r}
cyclistic_avgRide <- mean(cyclistic_date$ride_length)
print(cyclistic_avgRide)  
```  
#### Customer Type

##### Average ride_length:
```{r}
cyclistic_date %>% group_by( customer_type) %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```  
#### TYPE OF BIKE

##### Total rides by customer_type:
```{r}
cyclistic_date %>% group_by(customer_type, ride_type) %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```               
               
##### Average ride_length:
```{r}
cyclistic_date %>% group_by(ride_type) %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```

#### HOuR
##### Average ride_length by customer_type:
```{r}
cyclistic_date %>% group_by(hour, customer_type) %>% 
  summarise_at(vars(ride_length),
               list(time = mean)) %>% 
  print(n=48) #lets you view entire tibble
```
##### Average ride_length:
```{r}
cyclistic_date %>% group_by(hour) %>% 
  summarise_at(vars(ride_length),
               list(time = mean)) %>% 
  print(n=24) #lets you view entire tibble
```  
  
#### DAY OF THE WEEK

##### Average ride_length by customer_type:
```{r}
cyclistic_date %>% group_by(customer_type, day_of_week) %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```
##### Average ride_length: 
```{r}
cyclistic_date %>% group_by(day_of_week) %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```
#### DAY OF THE MONTH

##### Average ride_length by customer_type:
```{r}
cyclistic_date %>% group_by(day, customer_type) %>% 
  summarise_at(vars(ride_length),
               list(time = mean)) %>% 
  print(n=62)  #lets you view entire tibble
```
##### Average ride_length:
```{r}
cyclistic_date %>% group_by(day) %>% 
  summarise_at(vars(ride_length),
               list(time = mean)) %>% 
  print(n=31)  #lets you view entire tibble
```  
  
#### MONTH

##### Average ride_length by customer_type:
```{r}
cyclistic_date %>% group_by(month, customer_type) %>% 
  summarise_at(vars(ride_length),
               list(time = mean)) %>% 
  print(n=24)  #lets you view entire tibble
```
##### Average ride_length:
```{r}
cyclistic_date %>% group_by(month) %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```               
### SEASON

#### Spring:

##### Average ride length by customer_type:
```{r}
cyclistic_date %>% 
  group_by(customer_type) %>% 
  filter(season == "Spring") %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```

##### Average ride length:
```{r}
cyclistic_date %>% 
  filter(season == "Spring") %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```
#### SUMMER:

##### Average ride length by customer_type for summer: 
```{r}
cyclistic_date %>% 
  group_by(customer_type) %>% 
  filter(season == "Summer") %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```

##### Average ride length for summer:
```{r}
cyclistic_date %>% 
  filter(season == "Summer") %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```
#### Fall:

##### Average ride length by customer_type:
```{r}
cyclistic_date %>% 
  group_by(customer_type) %>% 
  filter(season == "Fall") %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```
##### Average ride length:
```{r}
cyclistic_date %>% 
  filter(season == "Fall") %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```
#### :

##### Average ride length by customer_type:
```{r}
cyclistic_date %>% 
  group_by(customer_type) %>% 
  filter(season == "Winter") %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```

##### Average ride length:
```{r}
cyclistic_date %>% 
  filter(season == "Winter") %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```
### All Season

#### Average ride length by customer_type:
```{r}
cyclistic_date %>% 
  group_by(season, customer_type) %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```
#### Average ride length: 
```{r}
cyclistic_date %>% 
  group_by(season) %>% 
  summarise_at(vars(ride_length),
               list(time = mean))
```
### Visualization:

#### Total trips by customer type Vs. Day_of_Week
```{r}

cyclistic_date %>%  
  group_by(customer_type, day_of_week) %>% 
  summarise(ride_length = n()) %>% 
  arrange(customer_type, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = ride_length, fill = customer_type)) +
  labs(title ="Total trips by customer type Vs. Day of the week") +
  geom_col(width=0.5, position = position_dodge(width=0.5)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
```

From the table and graph above, casual customers are most busy on Sundays followed by Saturdays, while members are most busy on later half of the week extending into the weekend. Interesting pattern to note though is the consistent trip numbers among members with less spread over entire week as compared to casual riders who don't seem to use the bikeshare services much during weekdays.

<br/>

#### Total trips by customer type Vs. Month

```{r}

cyclistic_date %>%  
  group_by(customer_type, month) %>% 
  summarise(ride_length = n()) %>% 
  arrange(customer_type, month)  %>% 
  ggplot(aes(x = month, y = ride_length, fill = customer_type)) +
  labs(title ="Total trips by Customer Type Vs. Month") +
  theme(axis.text.x = element_text(angle = 30)) +
  geom_col(width=0.5, position = position_dodge(width=0.5)) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE))
```

The data shows that the months of July, August, September and October are the most busy time of the year among both members and casual riders. This could be attributed to an external factor (eg. cold weather, major quality issue) that might have hindered with customer needs. 

<br/>


#### Visualizaton of average trip duration by customer type on each day of the week

```{r}
 cyclistic_date %>%  
  group_by(customer_type, day_of_week) %>% 
  summarise(average_trip_duration = mean(hour)) %>%
  ggplot(aes(x = day_of_week, y = average_trip_duration, fill = customer_type)) +
  geom_col(width=0.5, position = position_dodge(width=0.5)) + 
  labs(title ="Average trip duration by customer type Vs. Day of the week")

```

The average trip duration of a casual rider is little more than of a member.


<br/>

#### Visualizaton of average trip duration by customer type Vs. month
```{r}
cyclistic_date %>%  
  group_by(customer_type, month) %>% 
  summarise(average_trip_duration = mean(hour)) %>%
  ggplot(aes(x = month, y = average_trip_duration, fill = customer_type)) +
  geom_col(width=0.5, position = position_dodge(width=0.5)) + 
  labs(title ="Average trip duration by customer type Vs. Month") +
  theme(axis.text.x = element_text(angle = 30))
```

Average trip duration of member riders is anywhere between 10-30 minutes throughout the year, exception being February when it goes slightly over 20 minutes. However, there seems to be a distinct pattern when it comes to casual riders, whose average trip duration swings wildly from as low as ~25 minutes to more than an hour depending on time of the year. It is worth noting unusually long trip durations by casual riders in the month of February.

<br/>

#### Visualizaton of ride type Vs. number of trips by customer type
```{r}
cyclistic_date %>%
  group_by(ride_type, customer_type) %>%
  summarise(number_of_trips = n()) %>%  
  ggplot(aes(x= ride_type, y=number_of_trips, fill= customer_type))+
  geom_bar(stat='identity') +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  labs(title ="Ride type Vs. Number of trips")
```
<br/>

Classic bikes are predominantly used by members. Docked bikes are almost never used compared to others. Electric bikes are equally used by both members as well as casual riders. If docked bikes cost the highest among all 3 types
<br/>



## 6. Act

#### Important Findings

- Usage (based on trip duration) of bikes by casual riders is more than of member riders.

- Casual customers use bikeshare services more during weekends, while members use them consistently over the entire week.

- Average trip duration of casual riders is more than of member rider over any given day of the week cumulatively.

- Casual riders ride longer during first half of the year compared to the second half, while members clock relatively similar average trip duration month over month.

- Both Casual riders and Member rider prefer electric bikes and Classic bikes  while Docked bikes are popular among Casual members.

<br/>








